EEG2fMRI: Cross-Modal Synthesis for Functional Neuroimaging

### 2021-06-22 article/article.tex
Update article

\begin{abstract}
    Functional neuroimaging techniques are essential in advancing research in neuroscience. However, existing techniques require researchers to either compromise on spatial resolution, temporal resolution, cost, or portability. We attempt to model the  mapping between concurrent EEG and fMRI recordings using various machine learning architectures, in order to combine the complementary qualities of the two modalities. The trained model is then able to reconstruct simultaneous fMRI data solely from EEG data. Although initial results in a within-subject regime are promising, the limited quantity of multi-modal recordings currently available appears to be a major obstacle in generalizing to new subjects.
\end{abstract}

@@

The increased relevance of neural activity in cognitive science prompted the development of a large number of functional neuroimaging techniques. These techniques can be effectively described in terms of spatial resolution, temporal resolution (sampling rate), and portability. The most widely used functional neuroimaging techniques are EEG and fMRI. EEG scanners consist of arrays of electrodes placed directly on the user's scalp. They are extremely portable and have high temporal resolution. However, such scanners can only detect neural activity after first being filtered by the skull, leading to a major decrease in spatial resolution. Conversely, fMRI is based on electromagnetic waves which easily penetrate the skull. Therefore, it has high spatial resolution, but is lacking in most other aspects. Compared to EEG, it is hundreds of times more expensive and thousands of times slower. Moreover, it is virtually unmovable by its user.

Given the lack of techniques which individually exhibit high overall performance, there have been numerous attempts to integrate their complementary qualities by using multiple techniques simultaneously. EEG-informed fMRI, fMRI-informed EEG, and neurogenerative modelling are all methods of integrating data from simultaneous EEG and fMRI recordings. However, these methods require extensive access to both neuroimaging techniques involved and are limited to niched applications \cite{huster_methods_2012}.

@@

In addition to uni-modal synthesis, investigators have also proposed cross-modal synthesis for augmenting neuroimaging techniques. This consists of reconstructing data collected through a target modality based on data collected through a distinct source modality, by exploiting the inherent relation between them. In a setting where both source and target modalities are available, this task might have limited utility. However, the value of cross-modal synthesis comes from settings where only one modality is available. In such settings, existing data can be used to approximate data collected through the unavailable modality. Unfortunately, this approach has largely been used for augmenting structural neuroimaging, with limited analogous efforts for functional neuroimaging \cite{yi_generative_2019}.

@@

\section{Task}

@@

Second, if simultaneous multi-modal recordings are not available, but separate recordings of similar neural states with different modalities are available (i.e. two separate recording sessions based on the same task but using different techniques), semi-supervised learning can be used to model the mapping through a task-related signal. This is analogous to the task of machine translation, which is often based on sample pairing through a cross-lingual signal \cite{zhang_neural_2020}.

@@

\section{Data}

@@

The output part of a data point consists of a three-dimensional tensor of shape 53x63x52 which depicts the 3D "picture" of the subject's brain obtained through fMRI at the time roughly corresponding to the end of the EEG recording.

@@

The raw original dataset is structured as follows. There are 16 subjects (after excluding one whose data was corrupted), each undergoing 6 blocks of an oddball task. Each block contains 170 fMRI scans, obtained sequentially, at a frequency of 0.5 Hz (for a total of 340 seconds per block). Each block also contains EEG data, recorded constantly at 1000Hz for the entire duration of the block. From these longer sequences of data, we are extracting data points in the way described previously. However, we're discarding the first few fMRI scans obtained before 30 seconds of EEG managed to accumulate beforehand. This means that ~155 samples are derived per block, which translates to 930 per subject, which translates to 14880 in total.

@@

\section{Models}

A wide range of model architectures have been employed for learning the EEG-fMRI mapping. Initially, dense fully-connected networks were tasked with encoding the entire flattened EEG data of one sample into one embedding of lower dimensionality through a linear layer. This embedding would be further routed through subsequent linear layers. The last layer would contain a number of neurons equal to the number of voxels present in the fMRI scans, so that they could be trained to directly predict the voxel-level activity.

@@

>>>>

\begin{abstract}
    Functional neuroimaging techniques are essential in advancing research in neuroscience. However, existing techniques require researchers to either compromise on spatial resolution, temporal resolution, cost, or portability. We attempt to model the  mapping between concurrent EEG and fMRI recordings using various machine learning architectures (i.e. MLPs, CNNs, and transformers), in order to combine the complementary qualities of the two modalities. The trained model is then able to reconstruct simultaneous fMRI data solely from EEG data. Although initial results in a within-subject regime are promising, the limited quantity of multi-modal recordings currently available appears to be a major obstacle in generalizing to new subjects.
\end{abstract}

@@

The increased relevance of neural activity in cognitive science prompted the development of a large number of functional neuroimaging techniques. These techniques can be effectively described in terms of spatial resolution, temporal resolution (sampling rate), and portability. The most widely used functional neuroimaging techniques are EEG and fMRI. EEG scanners consist of arrays of electrodes placed directly on the user's scalp. They are extremely portable and have high temporal resolution. However, such scanners only measure the scalp surface, lacking the 3D representation of fMRI alternatives. Additionally, such scanners can only detect neural activity after first being filtered by the skull, leading to a major decrease in spatial resolution. Conversely, fMRI is based on electromagnetic waves which easily penetrate the skull. Therefore, it has high spatial resolution, but is lacking in most other aspects. It is orders of magnitude more expensive than EEG and typically has a temporal resolution of 2 seconds, in contrast with the much finer 1 millisecond of a standard EEG headset. Moreover, it is not portable.

Given the lack of techniques which individually exhibit high overall performance, there have been numerous attempts to integrate their complementary qualities by using multiple techniques simultaneously. EEG-informed fMRI, fMRI-informed EEG, and neurogenerative modelling are all methods of integrating data from simultaneous EEG and fMRI recordings \cite{huster_methods_2012}. However, these methods require extensive access to both neuroimaging techniques involved and are limited to niched applications.

@@

In addition to uni-modal synthesis, investigators have also proposed cross-modal synthesis for augmenting neuroimaging techniques \cite{yi_generative_2019}. This consists of reconstructing data collected through a target modality based on data collected through a distinct source modality, by exploiting the inherent relation between them. In a setting where both source and target modalities are available, this task might have limited utility. However, the value of cross-modal synthesis comes from settings where only one modality is available. In such settings, existing data can be used to approximate data collected through the unavailable modality. Unfortunately, this approach has largely been used for augmenting structural neuroimaging, with limited analogous efforts for functional neuroimaging.

@@

Several model architectures are considered to this end, namely fully-connected networks, convolutional neural networks, and transformers. All those architectures yield machine learning models which can be trained to approximate the EEG-fMRI mapping present in the dataset through typical statistical methods.

\section{Task Operationalization}

@@

Second, if simultaneous multi-modal recordings are not available, but separate recordings of similar neural states with different modalities are available (i.e. two separate recording sessions based on the same task but using different techniques), semi-supervised learning can be used to model the mapping through a task-related signal. This is analogous to the task of machine translation, which is often based on sample pairing through a cross-lingual signal \cite{zhang_neural_2020}. However, large qualities of data might be required to offset the weaker strength of the alignment signal (merely task-related, rather than based on the same exact neural activity).

@@

\section{Methods}

\subsection{Data}

@@

The output part of a data point consists of a three-dimensional tensor of shape 53x63x52 which depicts the 3D "picture" of the subject's brain obtained through fMRI at the time corresponding to the last two seconds of the EEG recording.

@@

The raw original dataset is structured as follows. There are 16 subjects (after excluding one whose data was corrupted), each undergoing 6 blocks of an oddball task. Each block contains 170 fMRI scans, obtained sequentially, at a frequency of 0.5 Hz (for a total of 340 seconds per block). Each block also contains EEG data, recorded constantly at 1000Hz for the entire duration of the block. From these longer sequences of data, we are extracting data points in the way described previously. However, we are discarding the first few fMRI scans obtained before 30 seconds of EEG managed to accumulate beforehand. This means that ~155 samples are derived per block, which translates to 930 per subject, which translates to 14880 in total.

@@

\subsection{Models}

A wide range of machine learning model architectures have been employed for optimizing the EEG-fMRI mapping. Initially, dense fully-connected networks were tasked with encoding the entire flattened EEG data of one sample into one embedding of lower dimensionality through a linear layer. This embedding would be further routed through subsequent linear layers. The last layer would contain a number of neurons equal to the number of voxels present in the fMRI scans, so that they could be trained to directly predict the voxel-level activity.

@@

\begin{table}
\begin{center}
        \begin{tabular}{|c | c|} 
            \hline
            Model Architecture & MSE \\
            \hline
            fully-connected network & 9.527e-4 \\ 
            convolutional neural network & 7.087e-4 \\ 
            \textbf{transformer} & \textbf{4.976e-4} \\ 
            \hline
           \end{tabular}
           \caption{Mean squared error of different models in a one-subject-left-out regime}
\end{center}
\end{table}
